Collaborative Multioutput Gaussian Processes: \\ Supplementary Material

### 2014-03-19 paper/inference.tex
3.1 done

%\subsection{Variational Inference in Sparse GPs Revisited}
%
%In this section we review inference in sparse GPs (as presented in Titsias).
%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
%

Inference in this model is done with variational inference.
In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points.
The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used in \citep{titsias2009variational,hensmangaussian} to derive variational inference for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (eq. \ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors.
This underlines the pivotal role of the inducing variables in this model as previously discussed.

\noindent
To find the best distribution $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergence terms are analytically tractable. To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}
The ELBO can be computed by first solving for the individual expectations 
$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ (see the supplementary material), which are then substituted into eq. \ref{eq:elbo} to give:
\begin{align}
\nonumber
&\log p(\y)
\ge \sum_{i,n}
\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{jnn} \\ \nonumber
         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
         - \frac{1}{2} \beta_i \sum_{j=1}^Q \trace w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
\nonumber
& - \sum_{j=1}^Q \bigg( \frac{1}{2} \log |\K_{jzz} \S_j^{-1}| + \frac{1}{2} \trace \K_{jzz}^{-1}  (\m_j \m_j^T + \S_j)\bigg) \\
\nonumber
& - \sum_{i=1}^P  \bigg( \frac{1}{2} \log |\K_{izz}(\Shi)^{-1}| + \frac{1}{2} \trace \K_{izz}^{-1} (\mhi (\mhi)^T + \Shi) \bigg) \\
  &\define \calL,
  \label{eq:L}
\end{align}
where $\K_{jzz} = k(\Zj, \Zj)$, $\K_{izz} = k(\Zhi, \Zhi)$, and with the help of auxiliary matrices $\A_j = k(\X,\Zj) \K_{jzz}^{-1}$ and $\Ahi = k(\X_i,\Zhi)\K_{izz}^{-1}$, 
\begin{align}
\tilde{\mu}_{in}
%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
\mat{\Lambda}_{jn}
%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
&= \A_j(n,:)^T \A_j(n,:)  \\
\mat{\Lambda}_{in}
%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
&= \Ahi(n,:)^T \Ahi(n,:) 
\end{align}
where $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$. 
Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \citep{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.

The novelty of the derived variational lower bound is that it decomposes across not only the outputs but also the inputs.
This enables the use of stochastic optimization methods which makes this model applicable to much larger datasets compared to the existing GP-based multioutput models.

\subsection{STOCHASTIC VARIATIONAL INFERENCE}
\newcommand{\oi}{\vec{o}_i}
% some notation

>>>>

%\subsection{Variational Inference in Sparse GPs Revisited}
%
%In this section we review inference in sparse GPs (as presented in Titsias).
%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
%
We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ across
multiple outputs using variational inference.
In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs. Section \ref{sec:svi}  takes advantage of this factorization to 
carry out stochastic variational inference which allows the model to scale to very large datasets.
Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.

\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as:
\begin{align}
p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
\end{align}
Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form:
\begin{align}
q(\g, \h, \u, \v | \y)
= p(\g|\u) p(\h|\v) q(\u,\v),
\end{align}
with
\begin{align}
q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
\end{align}
This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
to derive variational inference algorithms for the single output case.
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \ref{eq:gu} and \ref{eq:hv}), 
we only need to 
learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
the true posteriors.
This underlines the pivotal role of the inducing variables in this model as previously discussed.

\noindent
To find the best distribution $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
\begin{align}
%\nonumber
%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
%\nonumber
%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
\label{eq:elbo}
\end{align}
which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergence terms are analytically tractable. To compute the expected likelihood term in the ELBO we first see that
\begin{align}
\nonumber
\log \text{ } &p(\y | \u, \v)
% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
% \nonumber
\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
\newcommand{\Ahi}{\A^h_i}
\newcommand{\Zi}{\Z_i}

The ELBO can be computed by first solving for the individual expectations 
%$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ 
$ \Eb{\log p(y_{in} | \g_n, h_{in}) }$ over ${p(\g | \u) p(\h_i | \v_i)}$ and 
then substituting these into Equation \ref{eq:elbo}. (Please see 
the supplementary material for the details of the derivation.)  
Hence the resulting lower bound is given by:
\begin{align}
\nonumber
%&\log p(\y)
 & \calL = 
\sum_{i,n}
\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{jnn} \\ \nonumber
         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
         - \frac{1}{2} \beta_i \sum_{j=1}^Q \trace w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
\nonumber
& - \sum_{j=1}^Q \bigg( \frac{1}{2} \log |\K_{jzz} \S_j^{-1}| + \frac{1}{2} \trace \K_{jzz}^{-1}  \left( \m_j \m_j^T + \S_j \right)\bigg) \\
\nonumber
& - \sum_{i=1}^P  \bigg( \frac{1}{2} \log |\K_{izz}(\Shi)^{-1}| \\
& \quad \qquad + \frac{1}{2} \trace \K_{izz}^{-1} \left( \mhi (\mhi)^T   +   \Shi \right) \bigg)  \text{,}
  \label{eq:L}
\end{align}
%
where $\K_{jzz} = k(\Zj, \Zj)$, $\K_{izz} = k(\Zhi, \Zhi)$, and:
\begin{align}
\tilde{\mu}_{in}
%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \text{,}\\
\mat{\Lambda}_{jn}
%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
&= \A_j(n,:)^T \A_j(n,:)  \text{,}\\
\mat{\Lambda}_{in}
%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
&= \Ahi(n,:)^T \Ahi(n,:)  \text{,}
\end{align}
with $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$; 
$\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$; 
$\mu_{jn} = (\Mu_j)_n$;
 $\mu^h_{in} = (\Mu^h_i)_n$; 
and we have defined the auxiliary matrices $\A_j = k(\X,\Zj) \K_{jzz}^{-1}$ and $\Ahi = k(\X_i,\Zhi)\K_{izz}^{-1}$ and 
used $\A_j(n,:)$ to denote the $n$-th row vector of $\A_j$. 
Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \cite{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.

The novelty of the  variational lower bound in Equation  \eqref{eq:L} is that it decomposes across both inputs and outputs.
This enables the use of stochastic optimization methods, which 
allow the model to handle  ver large datasets for which existing GP-based multi-output models are simply 
impractical.  
%
\subsection{STOCHASTIC VARIATIONAL INFERENCE \label{sec:svi}}
\newcommand{\oi}{\vec{o}_i}
% some notation
